Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-06 paper-1/manuscript.tex
Rev F8

\begin{figure*}%
    \centering
    \includegraphics{fig8.pdf}
    \caption{}
    \label{fig8}
\end{figure*}
\textbf{Fig. 8} In the full model, STD reduces \dev{} non-target and increases \dev{} target responses relative to \msc{}.
\textbf{a} Average depression $D$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
\textbf{b} Relationship across neurons of the sample network between the contrast (\dev{} - \msc{}) in $D^A$ and the contrast in the target response, and associated linear regression.
\textbf{c} Median contrast (\dev{} - \msc{}) in $D^B$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
\textbf{d} Relationship across networks and stimuli between the median contrast in $D^B$ among the early portion and the median contrast in the response to B among the entire network, and associated linear regression.
\textbf{e} Network-wide median contrast in $D^A$, plotted against its Pearson correlation coefficient with the corresponding contrast in the target response, across networks and stimuli. The plot shows both that the negative correlation shown in panel b is robust (most networks are plotted with red markers, i.e. significantly negatively correlated), and that $\Delta D^A$ is less than 0 on average (see the inset histogram on the horizontal axis)

@@

To show that $D$ captures not only changes in $V_m$, by definition, but also in the resulting response, we correlated the contrast (\dev{} - \msc{}) in $D^A$ against the target response contrast across neurons of the sample network. Given the relatively large responses in the contrasted conditions, and the infrequent presentation of these trials, allowing causality to mostly flow from $D$ to response, rather than also in reverse, we are confident that this relationship roughly reflects a causal influence. As shown in Figure \ref{fig8}b, the relationship was clearly negative (Pearson's $\rho$ = -0.541, p = 2.28e-77; Wald test, n = 1000), indicating that $D$ reduces responses as expected.

Having gained a rough intuition of how individual neurons are affected by STD, we now turn to its effect on the network, working with statistics across networks and stimuli. First, in Figure \ref{fig8}c, we recapitulate the logic established in Figure \ref{fig6}: Given the spatial clustering of $D^B$, we surmised a similar local suppression effect as with TA. We found that, indeed, neurons responding early to stimulation in B showed greater $D^B$ in \dev{} than \msc{} (T = 7.26e+03, p = 1.01e-21; one-sided Wilcoxon, n = 120), while the late portion had slightly reduced $D^B$ (T = 2.24e+03, p = 0.000133). We can relate this directly to the non-target response contrasts in the early and late portions shown in Figure \ref{fig7}a, which followed a similar pattern, and take this to indicate that the increased presentation of B in \dev{} causes local suppression in terms of $D$. % Possible concern: The relationship dXnotA ~ dD^B is not clearly established.

To show that increased local suppression by STD reduces the non-target response, we correlated the median contrast in $D^B$ among the early portion of networks with the median contrast of the total response to B, echoing Figure \ref{fig6}e. As we show in Figure \ref{fig8}d, higher $\Delta D^B$ was clearly predictive of a lower network response ($\rho$ = -0.581, p = 1.81e-12; Wald test, n = 1000). This confirms again the integrity of the measure $D$, but also that STD-mediated local suppression played a role in reducing the non-target response in the \dev{} sequence.

Finally, given the very clear contrast in $D^A$ in the sample network, we considered the possibility that, in the context of the full model, STD could also evoke global suppression in the \msc{} sequence, contrary to our stated hypothesis. If so, we should expect to see less $D^A$ in \dev{} than \msc{}. To confirm this, we analysed the contrast (\dev{} - \msc{}) in $D^A$ across networks and stimuli, and found its network median to be robustly negative (T = 955, p = 1.23e-12; one-sided Wilcoxon, n = 120). In addition, as shown in Figure \ref{fig8}e, we correlated the contrast in $D^A$ with the target response contrast across neurons in each network, and found that weaker depression in \dev{} predicted a greater response in most networks (T = 724, p = 1.37e-14), though the median correlation coefficient $\rho$ was only -0.282. We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see Figure \ref{sfig8}). Taken together, we cautiously conclude that, while there is clear evidence for greater global suppression in \msc{} than \dev{} in terms of $D$, this difference appears to manifest only weakly in activity differences.

>>>>

\begin{figure*}%
    \centering
    \includegraphics{fig8.pdf}
    \caption{}
    \label{fig8}
\end{figure*}
\textbf{Fig. 8} In the full model, STD reduces \dev{} non-target and increases \dev{} target responses relative to \msc{}.
\textbf{a} Average depression $D$ in target (A) trials (top row) and B trials (bottom row) in the \dev{} (left) and \msc{} sequence (center), and their contrast (right) in the sample network. The color scale is shared across columns (and between \dev{} and \msc{}) for direct comparability.
\textbf{b} Relationship across neurons of the sample network between the contrast (\dev{} - \msc{}) in $D^A$ and the contrast in the target response, and associated linear regression.
\textbf{c} Network-wide median contrast in $D^A$, plotted against its Pearson correlation coefficient with the corresponding contrast in the target response, across networks and stimuli.
\textbf{d} Median contrast (\dev{} - \msc{}) in $D^B$ in the first 100 neurons to respond to B, and in the remaining, late portion of the network, across networks and stimuli.
\textbf{e} Relationship across networks and stimuli between the median contrast in $D^B$ among the early portion and the median contrast in the response to B among the entire network, and associated linear regression

@@

To show that $D$ captures not only changes in $V_m$, by definition, but also in the resulting response, we correlated the contrast (\dev{} - \msc{}) in $D^A$ against the target response contrast across neurons of the sample network. Notice that this mirrors the correlation shown in Figure \ref{fig4} between $\Delta V_{TA}$ before A and the response contrast. As shown in Figure \ref{fig8}b, the relationship was clearly negative (Pearson's $\rho$ = -0.541, p = 2.28e-77; Wald test, n = 1000), indicating that higher $D$ causes reduced responses as expected. We confirmed this relationship across networks and stimuli, see Figure \ref{fig8}c, and found that greater depression in \dev{} predicted a lower response in most networks (T = 724, p = 1.37e-14), though the median correlation coefficient $\rho$ was only -0.282. We note that calculating this same correlation in the model with STD only yields no significantly negative dependence (T = 3.99e+03, p = 0.828; see Figure \ref{sfig8}).

Having gained a rough intuition of how individual neurons are affected by STD, we now turn to its effect on the network, working with statistics across networks and stimuli. First, in Figure \ref{fig8}d, we recapitulate the logic established in Figure \ref{fig6}: Given the spatial clustering of $D^B$, we surmised a similar local suppression effect as with TA. We found that, indeed, neurons responding early to stimulation in B showed greater $D^B$ in \dev{} than \msc{} (T = 7.26e+03, p = 1.01e-21; one-sided Wilcoxon, n = 120), while the late portion had slightly reduced $D^B$ (T = 2.24e+03, p = 0.000133). We can relate this directly to the non-target response contrasts in the early and late portions shown in Figure \ref{fig7}a, which followed a similar pattern, and take this to indicate that the increased presentation of B in \dev{} causes local suppression in terms of $D$. % Possible concern: The relationship dXnotA ~ dD^B is not clearly established.

To show that increased local suppression by STD reduces the non-target response, we correlated the median contrast in $D^B$ among the early portion of networks with the median contrast of the total response to B, echoing Figure \ref{fig6}e. As we show in Figure \ref{fig8}e, higher $\Delta D^B$ was clearly predictive of a lower network response ($\rho$ = -0.581, p = 1.81e-12; Wald test, n = 1000). This confirms again the integrity of the measure $D$, but also that STD-mediated local suppression played a role in reducing the non-target response in the \dev{} sequence.

Finally, given the very clear contrast in $D^A$ in the sample network, we considered the possibility that, in the context of the full model, STD could also evoke global suppression in the \msc{} sequence, contrary to our stated hypothesis. If so, we should expect to see less $D^A$ in \dev{} than \msc{}. To confirm this, we analysed the contrast (\dev{} - \msc{}) in $D^A$ across networks and stimuli, and found its network median to be robustly negative (T = 955, p = 1.23e-12; one-sided Wilcoxon, n = 120), as can be seen in the inset histogram in Figure \ref{fig8}c. Taken together, we cautiously conclude that, while there is clear evidence for greater global suppression in \msc{} than \dev{} in terms of $D$, this difference appears to manifest only weakly in activity differences.
